---
title: "An Empirical Analysis of Baseball"
author: "Ramanpreet Chand"
date: "December 5, 2016"
output: 
  ioslides_presentation:
      #css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
setwd("C:/Workspace/baseball-project")


# Load Packages
library(RMySQL)
library(DBI)
library(dplyr)
library(ggplot2)
library(lubridate)
library(pryr)
library(data.table)

# Disable Scientific Notation
options(scipen = 999)

# Open MySQL connection
con <- dbConnect(RMySQL::MySQL(), user = 'root', password = 'pg_password', host = 'localhost', dbname = 'retrosheet')

# Close MySQL connection using on.exit()
#on.exit(dbDisconnect(con))

# Build Queries by decade
query.events <- dbSendQuery(con, "SELECT GAME_ID, YEAR_ID, BAT_ID, PIT_ID, BAT_TEAM_ID, FLD_TEAM_ID, H_CD, AB_FL, EVENT_CD, SH_FL, SF_FL, BATTEDBALL_CD, RBI_CT, DP_FL, TP_FL, PB_FL, BUNT_FL, BAT_HAND_CD, PIT_HAND_CD, PA_NEW_FL, ERR_CT, PO1_FLD_CD, PO2_FLD_CD, PO3_FLD_CD
                            FROM retrosheet.events WHERE YEAR_ID BETWEEN 1976 AND 2015")

# Store Queries in local data frames
data.events <- fetch(query.events, n = -1)

# Class Declarations
data.events$YEAR_ID <- factor(data.events$YEAR_ID)
data.events$AB_FL <- factor(data.events$AB_FL, labels = c('Not At Bat', 'At Bat'))
data.events$H_CD <- factor(data.events$H_CD, labels = c('No Hit', '1B', '2B', '3B', 'HR'))
data.events$EVENT_CD <- factor(data.events$EVENT_CD, labels = c('Out', 'K', 'SB', 'DI', 'CS', 'PK', 'WP', 'PB', 'BK', 'OA', 'FE', 'NIBB', 'IBB', 'HBP', 'XI', 'ROE', 'FC', '1B', '2B', '3B', 'HR'))
data.events$BATTEDBALL_CD <- factor(data.events$BATTEDBALL_CD, labels = c('Null', 'FB', 'GB', 'LD', 'PU'))
data.events$BUNT_FL <- factor(data.events$BUNT_FL)
data.events$SH_FL <- factor(data.events$SH_FL, labels = c('Not Awarded', 'Awarded'))
data.events$SF_FL <- factor(data.events$SF_FL, labels = c('Not Awarded', 'Awarded'))
data.events$DP_FL <- factor(data.events$DP_FL, labels = c('Not Awarded', 'Awarded'))
data.events$TP_FL <- factor(data.events$TP_FL, labels = c('Not Awarded', 'Awarded'))
data.events$PB_FL <- factor(data.events$PB_FL, labels = c('Not Awarded', 'Awarded'))
data.events$PA_NEW_FL <- factor(data.events$PA_NEW_FL, labels = c('Not Awarded', 'Awarded'))
data.events$PIT_HAND_CD <- factor(data.events$PIT_HAND_CD, labels = c('Both', 'Left', 'Right'))
data.events$BAT_HAND_CD <- factor(data.events$BAT_HAND_CD, labels = c('Unknown', 'Left', 'Right'))

# Declare new variables
data.events <- data.events %>%
    mutate(h = ifelse(H_CD == 'No Hit', 0, 1), 
           ab = ifelse(AB_FL == 'Not At Bat', 0, 1),
           bb = ifelse(EVENT_CD == 'NIBB', 1, 0), 
           hbp = ifelse(EVENT_CD == 'HBP', 1, 0),
           sf = ifelse(SF_FL == 'Not Awarded', 0, 1), 
           weightH = ifelse(EVENT_CD == '1B', 1, ifelse(EVENT_CD == '2B', 2, ifelse(EVENT_CD == '3B',3, ifelse(EVENT_CD == 'HR', 4, 0)))),
           cs = ifelse(EVENT_CD == 'CS', 1, 0), 
           ibb = ifelse(EVENT_CD == 'IBB', 1, 0),
           sh = ifelse(SH_FL == 'Not Awarded', 0, 1), 
           sb = ifelse(EVENT_CD == 'SB', 1, 0), 
           out = ifelse(EVENT_CD == 'Out', 1, 0),
           k = ifelse(EVENT_CD =='K', 1, 0), 
           bk = ifelse(EVENT_CD == 'BK', 1, 0),
           di = ifelse(EVENT_CD == 'DI', 1, 0),
           flyBall = ifelse(BATTEDBALL_CD == 'FB', 1, 0),
           groundBall = ifelse(BATTEDBALL_CD == 'GB', 1, 0),
           firstB = ifelse(EVENT_CD=='1B', 1, 0), 
           secB = ifelse(EVENT_CD=='2B',1,0), 
           thirdB = ifelse(EVENT_CD=='3B',1,0), 
           hr = ifelse(EVENT_CD=='HR', 1, 0),
           gidp = ifelse(DP_FL=='Not Awarded', 0, 1),
           tp = ifelse(TP_FL == 'Not Awarded', 0, 1),
           pb = ifelse(PB_FL == 'Not Awarded', 0, 1),
           pa = ifelse(PA_NEW_FL == 'Not Awarded', 0, 1),
           wp = ifelse(EVENT_CD == 'WP', 1, 0),
           roe = ifelse(EVENT_CD == "ROE", 1, 0))

fileID <- "player_ids.csv"
col <- c('BAT_ID', 'FIRST', 'LAST', 'DEBUT')
Player <- fread(fileID, data.table = FALSE, header = TRUE, showProgress = FALSE, na.strings = "NULL")
names(Player) <- col
# Substring
Player <- Player %>% mutate(DEBUT = substr(DEBUT, 0, 10)) %>% data.frame()
# Class Declaration
Player$DEBUT <- mdy(Player$DEBUT)

# Extra for Fielders and Pitchers' calculations
fileMas <- "lahmanDB/2015/baseballdatabank-master/core/Master.csv"
masVar <- c("playerID", "retroID")
master <- fread(fileMas, data.table = FALSE, showProgress = FALSE, na.strings = "NULL", select = masVar)

# Extra for Pitchers
file <- "lahmanDB/2015/baseballdatabank-master/core/Pitching.csv"
myVar <- c("playerID", "yearID", "W", "L", "G", "GS", "CG", "SHO", "SV", "IPouts", "ER", "BFP", "GF", "R")
Pitching.file <- fread(file, data.table = FALSE, showProgress = FALSE, na.strings = "NULL", select = myVar)
Pitching.file <- Pitching.file %>% filter(yearID >= 1976)

# Extra for Fielders 
fileField <- "lahmanDB/2015/baseballdatabank-master/core/Fielding.csv"
Fielding.file <- fread(fileField, data.table = FALSE, showProgress = FALSE, na.strings = "NULL", select=c("yearID", "teamID", "playerID", "POS", "InnOuts", "A"))
Fielding.file <- merge(Fielding.file, master, by="playerID")
Fielding.file <- Fielding.file %>% filter(yearID >= 1976)
colsF <- c("playerID", "YEAR_ID", "FLD_TEAM_ID", "POS", "InnOuts", "Assists", "BAT_ID")
names(Fielding.file) <- colsF
```

## An Introduction

A statistical analysis of baseball involves measuring the individual, team and league's performance based on two broad categories:

- Offense: players are in a position to score
- - Batting
- Defense : players are in a position to prevent 
- - Pitching
- - Fielding
  
Sabermetrics objectively measure player performance that contributes to a team's win or loss.

## Key Terminology - Batting

- At Bats : A batter's turn against a pitcher 
- Batting Average : Hits to At Bat ratio
- OBP * : On Base Percentage
- SLG * : Slugging measures the power of a batter
- OPS * : On Base Plus Slugging
- ISO * : Isolated Power
- BABIP * : Batting Average for Balls in Play
- Runs Created *
- - A Sabermetric used to estimate the number of runs an individual player contributes to a team.

## Key Terminology - Pitching

- WHIP * : Walks Plus Hits per Inning Pitched
- - Measures a pitchers effectiveness in preventin a batter from running
- LOB % * : Left On Base Percentage
- - Measures the ratio at which he leaves baserunners stranded by the end of the innings
- FIP * : Fielding Independent Pitching
- - Measure a pitcher's performance based on the events he can control
- DICE * : Defensive Independent Component ERA
- - Portrays a pitchers Earned Runs Average
  
## Key Terminology - Fielding

- TC : Total Chances a fielder has to handle a ball in play
- FP : Fielding Percentage
  - Measures effectiveness of Fielders
- RF : Range Factor
  - Measures a fielder's actual contribution to a team based on position
- DER : Defensive-Efficiency Ratio
  - Determines the overall effectiveness of a team's defense.

```{r echo=FALSE}
# Batting Statistics Summarized over Time
Batting.active <- data.events %>% group_by(YEAR_ID, BAT_ID, BAT_HAND_CD) %>% filter(sum(ab)>=250) %>%
  summarize(firstBase = sum(firstB), secondBase=sum(secB), thirdBase=sum(thirdB), HR=sum(hr), K=sum(k),
            H=sum(h), AB = sum(ab), BB=sum(bb), CS=sum(cs), GIDP=sum(gidp), HBP=sum(hbp), IBB=sum(ibb), PA=sum(pa), RBI=sum(RBI_CT), SF=sum(sf), SH=sum(sh), SB=sum(sb), TB=sum(weightH)) %>%
  as.data.frame()

# Class Declaration
Batting.active$YEAR_ID <- ymd(paste(Batting.active$YEAR_ID, "-01-01", sep=""))

# mutate new statistics for each roster
Batting.active <- Batting.active %>% group_by(YEAR_ID, BAT_ID, BAT_HAND_CD) %>%
  mutate(Avg = round(H/AB,3),
         OBP = round(100*((H + BB + HBP)/(AB + BB + HBP + SF)),2),
         SLG = round((TB/AB),3),
         OPS = round((sum(OBP)/100) + sum(SLG),3),
         ISO = (TB-H)/AB,
         RC = round((((H + BB + HBP - CS - GIDP) * (TB + 0.26 * (BB - IBB + HBP)) + (0.52 * (SH + SF + SB)))/(AB + BB + HBP + SH + SF)),3),
         BABIP = round(((H - HR)/(AB - K - HR + SF)),3)) %>% as.data.frame()

# Merge Datasets to Player
Batting.active <- merge(Batting.active, Player, by="BAT_ID")

Batting <- Batting.active %>% group_by(YEAR_ID) %>% summarize(Avg.avg = mean(Avg, na.rm=TRUE),
                                                              RC.avg = mean(RC, na.rm=TRUE),
                                                             OBP.avg = mean((OBP/100),na.rm=TRUE),
                                                             SLG.avg = mean(SLG, na.rm=TRUE),
                                                             OPS.avg = mean(OPS, na.rm=TRUE),
                                                             ISO.avg = mean(ISO, na.rm=TRUE),
                                                             BABIP.avg = mean(BABIP, na.rm=TRUE)) %>% as.data.frame()
# Career Leaders
Batting.Career <- Batting.active %>% group_by(BAT_ID, YEAR_ID) %>% 
  filter(BABIP>=quantile(BABIP, 0.25, na.rm=TRUE),
         BABIP<=quantile(BABIP, 0.75, na.rm=TRUE), 
         ISO>=quantile(ISO, 0.25, na.rm=TRUE),
         ISO<=quantile(ISO, 0.75, na.rm=TRUE),
         OPS>=quantile(OPS, 0.25, na.rm=TRUE),
         OPS<=quantile(OPS, 0.75, na.rm=TRUE),
         SLG>=quantile(SLG, 0.25, na.rm=TRUE),
         SLG<=quantile(SLG, 0.75, na.rm=TRUE),
         OBP>=quantile(OBP, 0.25, na.rm=TRUE),
         OBP<=quantile(OBP, 0.75, na.rm=TRUE),
         RC>=quantile(RC, 0.25, na.rm=TRUE),
         RC<=quantile(RC, 0.75, na.rm=TRUE)) %>% as.data.frame()

#Active Roster Qualifiers
Batting.Career.Active <- Batting.Career %>% group_by(BAT_ID, YEAR_ID) %>% filter(YEAR_ID >= "2008-01-01",
                                                                                 YEAR_ID <= "2014-01-01") %>% as.data.frame()

  #Average out the multi-appearing players

Batting.Career.Active.Player <- Batting.Career.Active %>% group_by(BAT_ID, BAT_HAND_CD) %>%
  summarize(RC.avg = mean(RC, na.rm=TRUE),
            OBP.avg = mean((OBP/100),na.rm=TRUE), 
            SLG.avg = mean(SLG, na.rm=TRUE), 
            OPS.avg = mean(OPS, na.rm=TRUE), 
            ISO.avg = mean(ISO, na.rm=TRUE), 
            BABIP.avg = mean(BABIP, na.rm=TRUE)) %>% as.data.frame()

Batting.Career.Active.Player <- merge(Batting.Career.Active.Player, Player, by="BAT_ID")

# Top 25 by Runs created avg
Batting.Career.Active.Player.Final <- Batting.Career.Active.Player %>% arrange(desc(RC.avg)) %>% head(25)
col <- c("BAT_ID", "BAT_HAND_CD", "RC", "OBP", "SLG", "OPS", "ISO", "BABIP", "First", "Last", "Debut")
names(Batting.Career.Active.Player.Final) <- col


# Match the active roster's actual performance to their expected performance
Batting.Career.Active.Future <- Batting.Career %>% group_by(BAT_ID, YEAR_ID) %>% filter(YEAR_ID == "2015-01-01") %>% as.data.frame()

#Average out the multi-appearing players

Batting.Career.Active.Player.Future <- Batting.Career.Active %>% group_by(BAT_ID, BAT_HAND_CD) %>%
  summarize(RC.avg = mean(RC, na.rm=TRUE),
            OBP.avg = mean((OBP/100),na.rm=TRUE), 
            SLG.avg = mean(SLG, na.rm=TRUE), 
            OPS.avg = mean(OPS, na.rm=TRUE), 
            ISO.avg = mean(ISO, na.rm=TRUE), 
            BABIP.avg = mean(BABIP, na.rm=TRUE)) %>% as.data.frame()

Batting.Career.Active.Player.Future <- merge(Batting.Career.Active.Player.Future, Player, by="BAT_ID")
```

## Linear Regression
```{r}
linearModel <- lm(formula = RC ~ BABIP + OBP + SLG + OPS + ISO + BAT_HAND_CD, data = Batting.active)
summary(linearModel)

# Predictions
Trout <- data.frame(Batting.Career.Active.Player.Final[1,2:8])
Cabrera <- data.frame(Batting.Career.Active.Player.Final[2,2:8])
Fielder <- data.frame(Batting.Career.Active.Player.Final[3,2:8])

Trout.model <- predict(linearModel, newdata = Trout)
Cabrera.model <- predict(linearModel, newdata = Cabrera)
Fielder.model <- predict(linearModel, newdata = Fielder)

Trout.Future <- Batting.Career.Active.Player.Future %>% filter(BAT_ID == "troum001")
Cabrera.Future <- Batting.Career.Active.Player.Future %>% filter(BAT_ID == "cabrm001")
Fielder.Future <- Batting.Career.Active.Player.Future %>% filter(BAT_ID == "fielp001")
```
